Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference

### 2017-06-26 workflow/workflow.Rmd
Edits to text.

---
title: "Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, UC Berkeley
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine
  - name: Kelly Street
    affilitation: Graduate Group in Biostatistics, UC Berkeley
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, UC Berkeley
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics and Department of Statistics
abstract: Abstracts should be up to 300 words and provide a succinct summary of the article. Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article. Citations should not be used in the abstract, and the use of abbreviations should be minimized.
keywords: single-cell, workflow, single-cell-genomics, bioconductor, rna-seq, gene-expression, statistical-models, r
bibliography: ref.bib
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
---

@@

Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells. To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle different problems like dimensionality reduction, clustering, and the inference of cell lineages and pseudotimes. While each individual method is useful to analyze scRNA-seq data, there is an increasing need for creating integrated workflows of these new statistical tools. This is all the more true that novel sequencing technologies are able to sequence more and more cells in a single sequencing run. For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mice brains.

@@

Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: dimensionality reduction adjusting for gene and cell-level covariates, robust and stable clustering using resampling-based sequential ensemble clustering, and inference of cell lineages and ordering of the cells by developmental progression. Along the workflow, we use a unique SummarizedExperiment object to create an easy to use sequence of steps, where for each steps a different R package is used.

@@

The workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium [@Fletcher2017]. The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through differentiation of globose basal cells (GBCs) which are the actively proliferating cells in the epithelium. When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs) which become activated to differentiate and reconstitute all major cell types in the epithelium.

The dataset we work with in this workflow was generated to study the differentitation from the HBCs stem cells to different cell types present in the epithelium. To map the developmental trajectories of the multiple cell lineages arising from the olfactory epithelium’s HBC stem cell, a scRNA-seq analysis was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. Then, the expression of each gene was quantified by counting the total number of reads mapped to each cell to get a cell by gene matrix of raw counts. Different lineages were assigned to each cell in the dataset using a set of statistical tools. Finally, results were confirmed experimentally using in vivo lineage tracing. Details on the data generation and steps of the analysis are available in [@Fletcher2017].

>>>>

---
title: "Bioconductor workflow for single-cell RNA sequencing: Dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, UC Berkeley
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine
  - name: Kelly Street
    affilitation: Graduate Group in Biostatistics, UC Berkeley
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, UC Berkeley
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics and Department of Statistics
abstract: Abstracts should be up to 300 words and provide a succinct summary of the article. Although the abstract should explain why the article might be interesting, care should be taken not to inappropriately over-emphasise the importance of the work described in the article. Citations should not be used in the abstract, and the use of abbreviations should be minimized.
keywords: single-cell, single-cell genomics, RNA-seq, gene expression, zero inflation, normalization, clustering, lineage inference, statistical models, workflow, R, Bioconductor
bibliography: ref.bib
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
---

@@

Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells. To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include dimensionality reduction, clustering, and the inference of cell lineages and pseudotimes. While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline. This is all the more true, with novel sequencing technologies that allow an increasing numer of cells to be sequenced in each run. For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.

@@

Here, we propose an integrated workflow for dowstream analysis, with the following three main steps: dimensionality reduction adjusting for gene and cell-level covariates, robust and stable clustering using resampling-based sequential ensemble clustering, and inference of cell lineages and ordering of the cells by developmental progression. Along the workflow, we use a unique SummarizedExperiment object to create an easy to use sequence of steps, where a different R package is used for each step.

@@

This workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium (OE) [@Fletcher2017]. The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium. When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs) which become activated to differentiate and reconstitute all major cell types in the epithelium.

The scRNA-seq dataset we work with was generated to study the differentitation of HBCs stem cells into different cell types present in the olfactory epithelium. To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. Then, the expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it. Cells were then assigned to different lineages using the statistical analysis pipeline presented in this workflow. Finally, results were validated experimentally using in vivo lineage tracing. Details on data generation and statistical methods are available in [@Fletcher2017] [SD: Add reference to Slingshot and ZINB-WaVE biorXiv manuscripts.].
